Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory

### 2018-03-02 Cars/paper.tex
paper updates

\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}

%TODO:
%Flesh out section 1
%Add authors, institutions, an abstract and bibtex references 
%Switch out manual references for Bibtex when everything else is done.

@@

\begin{itemize}
\item Self driving vehicles are becoming more common, eg Waymo, Tesla, GM.
\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the self driving vehicle.
\item While self driving vehicles reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg. Cameras and LiDAR), such as their position, velocity, and steering angle, see \citet{Woo2016a} and \citet{Tian2017}.
\item The self driving vehicle forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
\item These can assume the surrounding vehicles will maintain their current angle and velocity \citet{Gindele2010}
\citet{Waymo2017}, \citet{Bautista2017}, but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citet{Woo2016b}, \citet{Ding2013}, \citet{Woo2017}, \citet{Geng2017}.
\item This paper takes a statistical approach, fitting time series models to two extracted variables, velocity and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
\item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
\item A homogenous time series model shows a strong ability to predict the MAP trajectory
\item An important part of this modelling is to capture the uncertainty of car trajectories, as the self driving vehicle must react to the possibility of collision even when that possibility is relatively small.
\item There are many sources of heterogeneity in drivers, from their individual driving styles to the conditions of traffic directly around them.
\item Including heterogeneity in the model will allow the model to better adapt to drivers with more extreme behaviours, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
\item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
\item This model implies that any new vehicle observed by the self driving vehicle has its own parameter vector, related to previously observed vehicles through the shared prior.
\item The self driving vehicle must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.
\item Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
\item A three step strategy is employed to provide density forecasts for the future trajectory of newly observed vehicles:
\begin{enumerate}
\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution. This can occur before the self driving vehicle is on the road.
\item When a new vehicle is observed use the information in this shared distribution to model the associated new parameter vector. 
\item Update the parameter vector and all forecasts at short intervals.
\end{enumerate}
\item This paper uses standard Variational Bayes for the second step, and introduces an updating method for the third step; the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
\item This only uses the data observed in-between updates, allowing the previously observed history of that vehicle to be discarded.
\item This is important for two reasons:
\begin{enumerate}
\item The model can be updated faster as less data must be processed, so forecasts can be made using as much recent data as possible.
\item Self driving vehicles take in up to 1GB of data a second (several websites report this as quotes from different people eg. Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.
\end{enumerate}
\item Show that a homogenous time series model has strong predictive performance compared to constant models / neural networks while also providing density forecasts.
\item Build a heterogenous version of the homogenous model, show that the density forecasts improve. 
\item Show that the Updating VB required to implement the heterogenous model in real time is comparable to MCMC and standard VB.
\end{itemize}

\section{Data processing}
\label{sec:dataProcessing}
Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 Freeway in Los Angeles, California from 7:50 am to 8:35 am on June 15th, 2005. Data was collected by seven static cameras and processed by Cambridge Systematics Inc. The result is coordinates relative to the start of the road section for 6101 vehicles collected every 100 milliseconds. The average number of observations per vehicle is 672, corresponding to a 67.2 seconds of data.
Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which travel along the five major lanes or the enter from the entry/exit lane to the right, with one vehicle changing from Lane 2 to Lane 1. Vehicles that entered or exited midway through the freeway section are excluded in this paper. There is a curvature to the road, with bends occuring between 500 and 1000 feet and 1800 and 2000 feet.

\begin{figure}
\centering
\includegraphics[width=0.65\textwidth]{carPath}
\caption{The path of ten vehicles in the dataset. The freeway is split into five lanes, and an additional entry/exit lane to the far right, through which two vehicles from this sample have entered. There is curvature to the road slightly distored by the aspect ratio, with bends occuring around 500 and 1800 feet.}
\label{fig:rawData}
\end{figure}

Modern vehicles are capable of tracking lane marking, and thus can compensate for the curvature of the road as they calculate the location of surrounding vehicles in coordinates relative to their own position (Thuy and Leon 2010), rather than relative to a static point as provided by NGSIM. To compensate for this, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding cars, the coordinate system provided in the NGSIM dataset is transformed into a relative system where the road curvature is removed. This is facilitated by measuring the location of each vehicle relative to the estimate of the midpoint of each lane. \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to build a model of the lane midpoints for the entire road section.

The midpoints for each of the five lanes are estimated separately, using 100 randomly sampled vehicles each that did not change their lane over the observed period. For each sampled vehicle $i$ and time since entering the road $t$, the total distance travelled is calcualted as 
\begin{equation}
\label{distance}
d_{i, t} = \sqrt{(x_{i, t}) - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}
\end{equation}
From this distance metric, the average $x$ and $y$ coordinates at each given lane and for a given distance are each estimated by a pair of smoothing splines calculated with the R `stats' package. The smoothing splines are functions of the distance travelled and estimate the lane midpoints as  $\hat{x}_{i, t} = f(d_{i, t})$ and $\hat{y}_{i, t} = g(d_{i, t})$. Each remaining vehicle in the dataset uses the lane midpoint estimates $\{\hat{x}^*_{i, t}, \hat{y}^*_{i, t}\}$ fit from the spline model assocaiated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the devitation from the lane midpoint, by
\begin{align}
x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right) - \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
y^*_{i, t} &= d_{i, t}. \label{yRel}
\end{align}
The vehicles used to estimate the five spline models are excluded for any other purpose.
\begin{figure}
\centering
\includegraphics[width = 0.5\textwidth]{relCoord}
\caption{An example of the equation in (\ref{xRel}). The estimated trajectory of the lane midpoint is given by the solid curve, while a vehicle $i$ at a given point in time $t$ is given by the blue dot at $\{x_{i, t}, y_{i, t}\}$. This vehicle has travelled a distance equivalent to the red dot at $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$. The gradient of the midpoint at this point is given by the dashed line, which intercepts the x-axis at an angle of $\lambda = \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)$. The dotted line from the blue dot to the red intercepts the x-axis at an angle of $\psi = \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)$. The sign of the angle $\lambda - \psi$ determines whether the vehicle is to the left or right side of the lane, and thus has a negative or positive relative coordinate $x^*_{i, t}$. The absolute value of $x^*_{i, t}$ is the distance between the two dots.}
\label{fig:relCoord}
\end{figure}
Once relative coordinates are extracted, the relative position of any vehicle $i$ at any time $t$ is determined by its initial position $\{x^*_{i, 1}, y^*_{i, 1}\}$, as well as the history of that driver's inputs: velocity $\{v_{i, s} | s = 2, \dots, t\}$ and steering angle, $\{\delta_{i, s} | s = 2, \dots, t\}$ by the trigonometric relationship in Figure \ref{fig:motion}:
\begin{align}
x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
\end{align}
where $\delta_{i, t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{i, t} = x^*_{i, t-1}$.
From this relationship, the driver's inputs are be extracted by
\begin{align}
\delta_{i, t} &= \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right) \label{dEq} \\
v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
\end{align}
It is then useful to extract the acceleration, $a_{i, t}$ by the simple relation
\begin{equation}
\label{aEq}
a_{i, t} = v_{i, t} - v_{i, t-1}. 
\end{equation}
Data for vehicle $i$ up to time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $z_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$.
\begin{figure}
\centering
\includegraphics[width = 0.4\textwidth]{motion}
\caption{A vehicle at $x_{t-1}, y_{t-1}$ at time $t-1$ will be at $x_t, y_t$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
\label{fig:motion}
\end{figure}

\section{Trajectory Forecasts}
\label{sec:trajForecasts}

A crucial step in the safe operation of self driving vehicles is to forecast the future trajectory of surrounding vehicles so that collisions can be avoided. To produce these forecasts the data is split to place vehicles into two sets: a training set used to infer the posterior distribution of a parameter set before the self driving vehicle is on the road; and a test set used for forecasting the positions of vehicles observed while driving. The parameters associated with the vehicles in the test set may be inferred, but this inference is time constrained. If inference takes $M$ seconds, a forecast after observing a vehicle for $T$ seconds can only include data up to time $T - M$ to infer the posterior distribution of any parameters.

The remainder of Section \ref{sec:trajForecasts} will introduce the models used, while Sections \ref{sec:Inference} and \ref{sec:UVB} will build an inferential framework to facilitate inference of any additional parameters in constant time as $T$ increases to apply to online trajectory forecasting.

@@

The framework provided by (\ref{xEq}) and (\ref{yEq}) allows forecasts of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_T, y^*_T, v_T\}$, and forecasts of the future values of 
$\{a_{t}, \delta_{t} | t = T + 1, \dots, T+H\}$.

As data is observed every 100 milliseconds, large changes in $a_{i, t}$ and $\delta_{i, t}$ can occur over multiple time periods. This can be observed in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
\begin{equation}
A_{i, t} = \frac{a_{i, \tau_i + t}}{a_{i, \tau_i}}
\label{amax}
\end{equation}
where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$. On average, a vehicle will increase in acceleration for the 400 milliseconds before reaching maximum acceleration. The right panel plots the mean value of
\begin{equation}
D_{i, t} = \frac{\delta_{i, \omega_i + t} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
\label{dmax}
\end{equation}
where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that large absolute deviations in steering angle from $pi/2$ are associated with smaller deviations in the same direction in the the previous 300 milliseconds.

\begin{figure}[h]
\centering
\includegraphics[width = 0.95\textwidth]{dynamics}
\caption{Left: Mean values of acceleration as a percent of a vehicles maximum absolute acceleration for the 1000 milliseconds before, and 500 milliseconds after, a vehicle reaches its maximum absolute acceleration. Right: The analogous plot for the steering angle, with $\delta_t - \pi/2$ as a percentage of a vehicles maximum absolute value of $\delta_t - \pi/2$. The few observations before a variable reaches its maximum often have increasing values for the same variable.}
\label{fig:dynamics}
\end{figure}

These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed to model this, with
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
\end{align}
for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.

The parameter vector is transformed to $\mathbb{R}^{p + q + 2}$ and collected as 
\begin{equation*}
\label{thetaVec}
\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\},
\end{equation*}
where
\begin{equation}
\label{indPrior}
p(\theta) \sim \mathcal{N}\left(\mu, \Sigma \right)
\end{equation}
where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data is very small. $\Sigma = 10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix. This is referred to as the homogenous model throughout the paper.

@@

The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them. Heterogeneity is allowed by replacing the shared $\theta$ parameter vector with $N$ individual parameter vectors $\theta_i$ so that
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{aAR2} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{dAR2}
\end{align}
Using the prior in (\ref{indPrior}) for each $\theta_i$ formulates what is referred to as the independent model where each vehicle posterior is independent of all other vehicles, so $p(\theta_i | z_{1:N}, \theta_{j \neq i}) = p(\theta_i | z_i)$. An alternative approach is to augment the prior distribution of each $\theta_i$ with parameters $\beta$ in a hierarchical model, and infer the distribution of the $p(\theta_{1:N}, \beta | z_{1:N})$. The top panel of Figure \ref{fig:HierSingleKDE} plots a kernel density estimate of the posterior means of each $\theta_i, i = 1, \dots, 2000$ using the independent model with $p = q = 2$ at time $T = 500$. There is a wide range of values for the posterior mean of each variable, indicating heterogeneity across the drivers in the sample. The density of posterior means for both variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the density of posterior means of both $\phi_1$ and $\phi_2$ are multimodal. Due these features, a $K$ component mixture of multivariate normal distributions is chosen for $p(\theta_i | \beta)$, augmenting each driver with an auxiliary variable $k_i = 1, \dots, K$ such that 
\begin{equation}
\label{mixPrior}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
\end{equation}
and
\begin{equation}
k_i \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
\end{equation}
for each $i = 1, \dots, N$. Furthurmore, for each $j = 1, \dots, K$,  
\begin{align}
\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right)
\end{align}
The elements of $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$. Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$. The bottom panel of Figure \ref{fig:HierSingleKDE} plots the marginal distributions for $p(\theta_{2001} | \beta, z_{1:2000})$ for the hierarchical model with $p = q = 2$ and $K = 6$ using the same vehicles as the independent model in the top panel. The marginal distributions capture the skewness and multimodality present in the densities of posterior means from the independent model.

>>>>

\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}
\author{Nathaniel Tomasetti}
\date{}

%TODO: Add a conclusion

@@

Self driving vehicles are rapidly becoming more advanced, with development from large corporations such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020. These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery and selects a path to avoid any collisions. As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self driving vehicles. However self driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
\\

Sophisticated Deep Neural Networks are employed to extract the position, velocity, and steering angle of surrounding traffic from readings provided by sensors such as cameras and Light Detection and Ranging (LiDAR); see \citet{Woo2016a} and \citet{Tian2017} for details. From this data, estimates of the future trajectory of surrounding vehicles can be produced. Forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
\\

This paper takes a statistical approach to the trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset. It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability collisions to be detected. The time series models used can allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them.
Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
\\

Incorporating heterogeneity into forecasts requires the behaviour of each driver of interest to be inferred as vehicles are encountered while driving. This inference is facilitated by first obtaining the posterior distribution for a large number of vehicles before the self driving vehicle is on the road, from which the prior distribution for additional vehicles can be constructed. Once the self driving vehicle is on the road, online inference is made possible by the introduction of Updating Variational Bayes (UVB), which approximates Bayesian updating so that inference may be periodically updated in constant time as additional data of surrounding vehicles is made available. This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used an steps required extract acceleration and steering angle for each vehicle, and Section \ref{sec:models} introduces several time series models for these variables. Section \ref{sec:Inference} reviews Bayesian inference, while Section \ref{sec:UVB} extends Variational Bayes to the online updating setting. Finally Section \ref{sec:eval} contains empirical results and Section \ref{sec:conc} concludes.

\section{Data processing}
\label{sec:dataProcessing}
Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005. Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section. 
Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1. Vehicles that entered or exited midway through the freeway section, are excluded in this paper. There is a curvature to the road occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.
\\

\begin{figure}
\centering
\includegraphics[width=0.65\textwidth]{carPath}
\caption{The path of ten vehicles in the dataset, with each black line representing a unique vehichle. This section of US 101 is split into five main lanes, with an additional entry/exit lane to the far right, through which two of the vehicles have entered. There is curvature to the road (which is slightly distorted by the aspect ratio) with bends occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.}
\label{fig:rawData}
\end{figure}

Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature. To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates. This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane. \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway. Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
\\

Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from time $t-1$ to time $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
\begin{align}
x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
\end{align}
Note that when $\delta_{i, t} \pm \pi/2$ then car $i$ takes a position parallel to the centre line, and hence $x^*_{i, t} = x^*_{i, t-1}$.
\\

From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs from driver $i$ are calculated via
\begin{align}
\delta_{i, t} &= 
     \begin{cases}
       \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right)  &\quad\text{if }x^*_{i, t} \neq x^*_{i, t-1} \\
       \frac{\pi}{2} &\quad\text{if } y^*_{i, t} > y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       -\frac{\pi}{2} &\quad\text{if } y^*_{i, t} < y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       \delta_{i, t-1} &\quad\text{otherwise} \\ 
     \end{cases} \label{dEq} \\
v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
\end{align}
In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=2,3,...,T\}$, according to
\begin{equation}
\label{aEq}
a_{i, t} = v_{i, t} - v_{i, t-1}. 
\end{equation}
\\

Data for vehicle $i$ up to time $T_i$ is defined as $z_{i, 1:T_i} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T_i\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$. 3373 vehicles remain in the dataset after the above exclusion criteria are applied. The data into two sets of vehicles: a training set of 2000 vehicles encountered before driving, and a test set of 1373 vehicles encountered while driving. 
\begin{figure}
\centering
\includegraphics[width = 0.4\textwidth]{motion}
\caption{A vehicle at coordinate $\{x^*_{t-1}, y^*_{t-1}\}$ at time $t-1$ will be at $\{x^*_t, y^*_t\}$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
\label{fig:motion}
\end{figure}

\section{Models for Trajectory Forecasting}
\label{sec:models}

Three similar models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles. These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent model and the hierarchical model. The independent model does not incorporate information from the training set of vehicles into inference for the test set, while the hierarchical model allows information to be shared between vehicles; it learns the range of behaviours possible from the training set and incorporates this into inference about the test set.

@@

The framework provided by (\ref{xEq}) and (\ref{yEq}) allows forecasts of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_T, y^*_T, v_T\}$, and forecasts of the future values of 
$\{a_{t}, \delta_{t} | t = T + 1, \dots, T+H\}$.  
\\

\iffalse % Partial Autocorrelation plots are a better way to show dynamic behaviour in the data.
This is highlighted in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
\begin{equation}
A_{i, t} = \frac{a_{i, \tau_i + t}}{a_{i, \tau_i}}
\label{amax}
\end{equation}
where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$. On average, large vehicle acceleration or deceleration is preceeded by an increase in acceleration in the same direction for 400 milliseconds. The right panel plots the mean value of
\begin{equation}
D_{i, t} = \frac{\delta_{i, \omega_i + t} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
\label{dmax}
\end{equation}
where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that steering angle deviation from $\pi/2$ are associated with smaller deviations in the same direction in during the previous 300 milliseconds. 
\begin{figure}[ht]
\centering
\includegraphics[width = 0.95\textwidth]{dynamics}
\caption{Left: Mean values of acceleration as a percent of a vehicles maximum absolute acceleration for the 1000 milliseconds before, and 500 milliseconds after, a vehicle reaches its maximum absolute acceleration. Right: The analogous plot for the steering angle, where deviations in angle from $\pi/2$ are plotted as a percentage of a vehicles maximum absolute value of deviation. The few observations before either variable reaches its maximum often have increased values in the same direction.}
\label{fig:dynamics}
\end{figure}
\fi

NGSIM vehicle data is observed every 100 milliseconds so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple sequential observations. This is highlighted in Figure \ref{fig:pacf}, which plots the partial autocorrelation function, which measures the correlation between two time periods $t$ and $t + k$, conditioned on $t + 1, \dots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for a selection of vehicles in the sample. Large spikes in the partial autocorrelation plots, such as those at the first and second lag for most vehicles, demonstrate that the value of either $a_t$ or $\delta_t$ is strongly correlated with the previous time periods. 
\\

\begin{figure}
\centering
\includegraphics[width = 0.95\textwidth]{pacf}
\caption{Top: Partial Autocorrelation plots for acceleration, $a$, for a selection of vehicles. Bottom: Partial Autocorrelation plots for steering angle, $\delta$ for the same vehicles. Each spike for vehicle $i$ at lag $k$ in the top panel indicates that $a_{i, t-k}$ is correlated with $a_{i, t}$, similarly spikes in the bottom panel indicate that $\delta_{i, t-k}$ is correlated with $\delta_{i, t}$. Note that different vehicles do not have the same dynamic behaviour in the partial autocorrelations.}
\label{fig:pacf}
\end{figure}

These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed with
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
\end{align}
for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.
\\

The parameters are transformed to $\mathbb{R}^{p + q + 2}$ and collected as the vector
\begin{equation*}
\label{thetaVec}
\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\},
\end{equation*}
with the prior
\begin{equation}
\label{indPrior}
\theta \sim \mathcal{N}\left(\mu, \Sigma \right),
\end{equation}
where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data at the 100 millisecond time-scale are very small. The prior variance $\Sigma$ is set to $10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix. This is referred to as the homogenous model throughout the paper.

@@

The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them. Heterogeneity is allowed by replacing the shared $\theta$ parameter vector with $N$ individual parameter vectors $\theta_i$ so that
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{aAR2} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{dAR2}
\end{align}
Use of the prior (\ref{indPrior}) for each $\theta_i$ formulates what is referred to as the independent model; where each individual $\theta_i$ posterior is independent of all other, i.e. $p(\theta_i | \textbf{z}_{1:N}, \theta_{j \neq i}) = p(\theta_i | z_i)$. 
\\

An alternative approach is to augment the prior distribution of each $\theta_i$ with parameters $\beta$ in a hierarchical model, and infer the distribution of $p(\theta_{1:N}, \beta | \textbf{z}_{1:N})$. The top panel of Figure \ref{fig:HierSingleKDE} plots a kernel density estimate of the posterior means of each $\theta_i, i = 1, \dots, 2000$, using the independent model with $p = q = 2$ at time $T = 500$. There is a wide range of values for the posterior mean of each variable, indicating heterogeneity across the drivers in the sample. The density of posterior means for both variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the density of posterior means of both $\phi_1$ and $\phi_2$ are multimodal. Due these features a $K$ component mixture of multivariate normal distributions is chosen for $p(\theta_i | \beta)$, augmenting each driver with an auxiliary variable $k_i = 1, \dots, K$ such that 
\begin{equation}
\label{mixPrior}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
\end{equation}
and
\begin{equation}
k_i \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
\end{equation}
for each $i = 1, \dots, N$. Furthurmore, for each $j = 1, \dots, K$,  
\begin{align}
\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right).
\end{align}
\\

The hyperparameters $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$. Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$. The bottom panel of Figure \ref{fig:HierSingleKDE} plots the marginal distributions for $p(\theta_{2001} | \beta, z_{1:2000})$ for the hierarchical model with $p = q = 2$ and $K = 6$ using the same data as the independent model. The hierarchical marginal distributions have captured the skewness and multimodality present in the densities in the top panel.
